Bayesian MCMC diagnostics and summarisation using Tracer 1.6

### 2017-07-28 manuscript/tracer16.tex

Bayesian inference of phylogeny \citep{huelsenbeck2001bayesian, felsenstein2004inferring,yang2006computational} has been a popular approach to reconstruct the evolutionary history and discover the evolutionary relationships among organisms that could come from genes, genomes, individuals or species.
The statistical framework is based on the posterior probability of phylogenetic trees given the data, and currently takes the advantage of Markov chain Monte Carlo (MCMC) to promote the computational speed of Bayesian estimation to reach an approximated optimisation.
The posterior summaries, as well as as scientifically relevant marginalisation, %TODO
are therefore critically important for understanding a Bayesian phylogenetic analysis.
%However, the full posterior distribution is high-dimensional and often difficult to visualise. 
We then develop a software package Tracer (version 1.6) for this purpose, which analyses Bayesian MCMC log files containing the list of parameters whose values sampled from the chain. 
It can be used to analyse runs of BEAST \citep{drummond2007beast,drummond2012bayesian}, BEAST2 \citep{bouckaert2014beast2}, MrBayes \citep{ronquist2012mrbayes}, RevBayes \citep{hohna2016revbayes}, LAMARC \citep{kuhner2006lamarc}, Migrate \citep{beerli2006comparison},  and possibly other MCMC programs.
In addition, Tracer will be able to redraw the distribution of effective population sizes over time if a demographic model was involved in the analysis.

@@

Tracer is able to analyse either a single MCMC log or combine the shared traces from multiple logs as designed in Figure~\ref{fig:tracer}. 
The list of traces in each log are created from the sampling parameters logged during a MCMC process, which include their statistical summaries from their values at all states logged. 
The effective sample size (ESS) is one of important measurements of a sampled continuous parameter, which tells the number of effectively independent draws from the posterior distribution that the Markov chain is equivalent to.

Tracer handles three trace types, also known as variable types \citep{mendenhall2012introduction}: real, integer, and categorical.
The real trace is used to handle quantitative continuous variables, which is the most common type in logged parameters.
The integer is dealing with quantitative discrete variables, and the categorical trace is for qualitative categorical variables. 
% check about definition 
These three types can be automatically assigned when the log file is imported. But they can also be changed later.

The multiple traces can be selected as shown in Figure~\ref{fig:multitrace}, which will overlay the plots for the different traces allowing comparisons to be made. The multiple trace files can be similarly selected as well to compare different runs. If multiple trace files have the same trace names then a ``Combined'' trace will automatically appear, which can be selected as well as the individual trace files.

@@

\item Density - this shows the Bayesian posterior density plot for the selected trace(s), if they are continuous, the kernel density estimates will be available, for example, in Figure~\ref{fig:4tabs}.b.

@@

Tracer also provides demographic reconstruction resulting in a graphical plot to estimate the distribution of effective population sizes over time, given a MCMC log from a Bayesian framework or additionally with the tree log depending on the model.
For example, ``Demographic Reconstruction" performs reconstructions of the distribution for a number of models, such as constant size \citep{drummond2002estimating}, exponential growth \citep{drummond2002estimating}, and logistic growth (?). %citation?
Besides ``Bayesian Skyline Reconstruction'' reconstructs a generalised Bayesian skyline plot (e.g. Figure~\ref{fig:flu}.a) describing demographic history under canonical scenarios \citep{drummond2005bayesian},
``GMRF Skyride Reconstruction'' and ``SkyGrid Reconstruction'' are respectively available for a temporal smoothing method using the Gaussian Markov random field (GMRF) model  \citep{minin2008smooth} and a generalisation of the GMRF model \citep{gill2012improving}.
``Lineages Through Time'' (?) plots quantiles of the number of lineages against time, which describes the change of diversification of all organisms over time. %citation?
All these models are available in BEAST \citep{drummond2007beast}.

%TODO better to re-write
Tracer recently offers a fairly general solution of looking at conditional posterior distributions. This, specifically, supports for BSSVS forms of model averaging, in which some parameters are only in the likelihood when their sub-model is ``indicated" by some indicator function that is usually a discrete, integer or boolean variable. In that case the posterior of the parameter should not include the states when it was sampled only in the prior because the sub-model it belonged to was ``turned off". This is relevant for EBSP, Random Local Clocks model, microsatellite model averaging and relaxed clock model averaging methods all from my group in last couple of years. It is also relevant for BSSVS in phylogeography as well depending on how the state is logged.

@@

Figure~\ref{fig:4tabs} demonstrates a snapshot of posterior summaries in Tracer for a BEAST \citep{drummond2007beast} run created from a 898 bp alignment of sequences of 12 species of primates (http://beast.bio.ed.ac.uk/tutorials?). %data citation
%using HKY + $\Gamma$ model, uncorrelated lognormal relaxed molecular clock, and Yule prior on the tree.
In this analysis, we furthermore specify three calibrations on the phylogenetic tree to estimate time to the most recent common ancestor (TMRCA) based on our prior fossil knowledge.

The first TMRCA prior, called ``ingroup'', includes all species except the \textit{Lemur} which will form the outgroup in the tree. %monophyletic
The second group is ``Human-Chimp'' that contains only \textit{Homo sapiens} and \textit{Pan}. 
The last consists of everything except \textit{Lemur}, \textit{Saimiri} and \textit{Tarsius} to create the hominoid-cercopithecoid split, which is named as ``HomiCerco''. 
Since we know the most recent common ancestor of humans and chimps may live on about $5-7$ million years ago, a normal distribution centered at 6 million years with a standard deviation of 0.5 million years is assigned to the ``Human-Chimp'' prior. 
Another normal distribution of $24 \pm 0.5$ (million years) is likewise defined for ``HomiCerco''.

@@

Selecting the ``treeModel.rootHeight'' parameter gives the marginal posterior distribution of the age of the root of entire tree. In the example in Figure~\ref{fig:4tabs}.a, we can conclude the root of the tree is most likely between about 39.01 and 63.67 million years old with the mean of 50.48 million years.

According to the kernel density estimation of the two TMRCA parameters (``Human-Chimp'' in blue and ``HomiCerco'' in yellow) in Figure~\ref{fig:4tabs}.b, these two parameters that we used to calibrate the tree explicitly have posterior distributions very similar to the prior distributions that we specified.

The marginal posterior densities of ``treeModel.rootHeight'' and ``ingroup'' are nearly overlapped each other. If we switch to Figure~\ref{fig:4tabs}.c, their joint-marginal distribution plot is almost lined up together, which indicates ??. But this isn't an issue here, because the ``ingroup'' only excludes one outgroup species from everyone else. %TODO

@@

\subsection*{Demographic reconstruction}

The second analysis uses an 1698bp alignment of 21 sequences from the Influenza A virus (?). %data citation
The piecewise-constant Bayesian skyline is selected for the tree prior, where the number of groups is reduced to 5 to avoid over-parameterize the model.

\begin{figure}[ht]
%\includegraphics[width=.5\textwidth]{./figures/fluBSP.pdf}  
\subfigure[Bayesian Skyline]{\includegraphics[width = .23\textwidth]{./figures/fluBSP.pdf}}
\subfigure[Lineages Through Time]{\includegraphics[width = .23\textwidth]{./figures/fluLTT.pdf}}
\caption{The plot (a) performs Bayesian Skyline reconstruction to estimate population sizes over time and (b) Lineages Through Time , from ``Flu'' data, where time starts from year 1997 at the 0 position and the unit is year.}
\label{fig:flu}
\end{figure}

@@

\begin{figure}[H]
\includegraphics[width=.5\textwidth]{./figures/comp-95HPD.pdf}  
\caption{The comparison of  95\% HPD intervals from multi-trace}
\label{fig:comp95HPD}
\end{figure}

\begin{figure}[H]
\includegraphics[width=.5\textwidth]{./figures/jointPrInt.png}  
\caption{The Joint probability table of two integer traces}
\label{fig:int:jointpr}
\end{figure}

>>>>

Bayesian inference of phylogeny \citep{huelsenbeck2001bayesian, felsenstein2004inferring,yang2006computational} has been a popular approach to reconstruct the evolutionary history and discover the evolutionary relationships among organisms that could come from genes, genomes, individuals or species.
The statistical framework is based on the posterior probability of phylogenetic trees given the data, and currently takes the advantage of Markov chain Monte Carlo (MCMC) to promote the computational speed of Bayesian estimation to reach an approximated optimisation.
The posterior summaries, as well as as scientifically relevant marginalisation, %TODO
are therefore critically important for understanding a Bayesian phylogenetic analysis.
%However, the full posterior distribution is high-dimensional and often difficult to visualise. 
We then develop a software package Tracer (version 1.6) for this purpose, which analyses Bayesian MCMC log files containing the list of parameters whose values sampled from the chain. 
It can be used to analyse runs of BEAST \citep{drummond2007beast,drummond2012bayesian}, BEAST2 \citep{bouckaert2014beast2}, MrBayes \citep{ronquist2012mrbayes}, RevBayes \citep{hohna2016revbayes}, LAMARC \citep{kuhner2006lamarc}, Migrate \citep{beerli2006comparison},  and possibly other MCMC programs.
In addition, Tracer's demographic reconstruction functions are utilised to redraw the distribution of effective population sizes over time if an available demographic model was involved in the analysis.

@@

Tracer is able to analyse either a single MCMC log or combine the shared traces from multiple logs as designed in Figure~\ref{fig:tracer}. 
The list of traces in each log are created from the sampling parameters logged during a MCMC process, which include their statistical summaries from their values at all states logged. 
The effective sample size (ESS) is an important measurement designed for a sampled parameter having continuous values, which tells the number of effectively independent draws from the posterior distribution that the Markov chain is equivalent to.

Tracer handles three trace types, also known as variable types \citep{mendenhall2012introduction}: real, integer, and categorical.
The real trace is used to handle quantitative continuous variables, which is the most common type in logged parameters.
The integer is dealing with quantitative discrete variables, and the categorical trace is for qualitative categorical variables. 
% check above definition 
%These three types can be automatically assigned when the log file is imported. But they can also be changed later.

The multiple traces can be selected as shown in Figure~\ref{fig:multitrace}, which will overlay the plots for the different traces allowing comparisons to be made. The multiple trace files can be similarly selected as well to compare different runs. If multiple trace files have the same trace names,  then a ``Combined'' trace will automatically appear, which can be selected as well as the individual trace files.

@@

\item Density - this shows the Bayesian posterior density plot for the selected trace(s), if they are continuous, the kernel density estimates will be available for continuous traces, for example, in Figure~\ref{fig:4tabs}.b.

@@

Tracer also provides demographic reconstruction resulting in a graphical plot, often applyed to reconstruct epidemic dynamics, given a MCMC log from a Bayesian framework or additionally with the tree log depending on the model.
The currently available models are constant size \citep{drummond2002estimating}, exponential growth \citep{drummond2002estimating}, logistic growth (?), Bayesian coalescent skyline \citep{drummond2005bayesian}, GMRF Bayesian skyride \citep{minin2008smooth}, and Bayesian coalescent skygrid \citep{gill2012improving}.
%For example, ``Demographic Reconstruction" performs reconstructions of the distribution for a number of models, such as constant size \citep{drummond2002estimating}, exponential growth \citep{drummond2002estimating}, and logistic growth (?). %citation?
%Besides ``Bayesian Skyline Reconstruction'' reconstructs a generalised Bayesian coalescent skyline plot (e.g. Figure~\ref{fig:flu}.a) describing demographic history under canonical scenarios \citep{drummond2005bayesian},
%``GMRF Skyride Reconstruction'' and ``SkyGrid Reconstruction'' are respectively available for a temporal smoothing method using the Gaussian Markov random field (GMRF) model  \citep{minin2008smooth} and a generalisation of the GMRF model \citep{gill2012improving}.
Besides these, ``Lineages Through Time'' (?) is also available to plot the quantiles of the number of lineages against time, which describes the change of diversification of all organisms over time. %citation?
%All these models are available in BEAST \citep{drummond2007beast}.

%TODO is it OK ?
Tracer recently offers a fairly general solution of looking at conditional posterior distributions. This, specifically, supports for Bayesian stochastic search variable selection (BSSVS) forms of model averaging, in which some parameters are only in the likelihood when their sub-model is indicated by an indicator function usually assigned to a discrete, integer or boolean variable. In that case, the posterior of the parameter should not include the states when it was sampled only in the prior, because the sub-model it belonged to was ``turned off". This is relevant for extended Bayesian skyline plot, random local clock model, microsatellite model averaging, and relaxed clock model averaging methods, etc. It is also relevant for BSSVS in phylogeography as well depending on how the state is logged.

@@

Figure~\ref{fig:4tabs} demonstrates a snapshot of posterior summaries in Tracer for a BEAST \citep{drummond2007beast} run created from a 898 bp alignment of sequences of 12 species of primates (http://beast.bio.ed.ac.uk/tutorials?). %data citation

%using HKY + $\Gamma_4$ model, uncorrelated lognormal relaxed molecular clock, and Yule prior on the tree.
In this analysis, we furthermore specified three calibrations on the phylogenetic tree to estimate time to the most recent common ancestor (TMRCA) based on our prior fossil knowledge.

The first TMRCA prior, called ``ingroup'', includes all species except the \textit{Lemur} which will form the outgroup in the tree. %monophyletic
The second group is ``Human-Chimp'' that contains only \textit{Homo sapiens} and \textit{Pan}. 
The last consists of everything except \textit{Lemur}, \textit{Saimiri} and \textit{Tarsius} to create the hominoid-cercopithecoid split, which is named as ``HomiCerco''. 
We know the most recent common ancestor of humans and chimps may live about $5-7$ million years ago, a normal distribution centered at 6 million years with a standard deviation of 0.5 million years is assigned to the ``Human-Chimp'' prior. 
Another normal distribution of $24 \pm 0.5$ (million years) is likewise defined for ``HomiCerco''.

@@

After loading the log, selecting the ``treeModel.rootHeight'' parameter gives the marginal posterior distribution of the age of the root of entire tree. The blue bars in Figure~\ref{fig:4tabs}.a present the most compact interval on the root height of the tree that contains 95\% of the posterior probability, referring to the period between about 39.01 and 63.67 million years old with the mean of 50.48 million years.
%TODO rate is faster than Yang's book r_ape = 5.4 * 10^-10 per year, r_monkey = 8.0, r_root = 6.6 
Correspondingly selecting ``meanRate'' provides the rate of evolution averaged over the whole tree, which has the mean of 0.0129 substitutions per site per million years with 95\% HPD Interval between 0.0101 and 0.0159.

Figure~\ref{fig:4tabs}.b shows the kernel density estimations of these four parameters. The two TMRCA parameters, ``Human-Chimp'' in blue and ``HomiCerco'' in yellow, that we used to calibrate the tree explicitly have posterior distributions very similar to the prior distributions that we specified.

The marginal posterior densities of ``treeModel.rootHeight'' and ``ingroup'' are nearly overlapped each other. When we switch to their joint-marginal distribution plot in Figure~\ref{fig:4tabs}.c, the samples are almost lined up together, which indicates they are associated. 
But this isn't an issue here, because the ``ingroup'' only excludes one outgroup species from all taxa. 
%TODO how to explain this problem further?

@@

\subsection*{Non-parametric coalescent estimator}

In the second example, there are two analyses conducted using a H3N2 data set that is the subset of a comprehensive influenza A virus data set spanning several epidemic seasons in the New York state. The data contains a  ?bp alignment of 165 Hemagglutinin genes collected between about 1999 and 2004 \citep{rambaut2008genomic}, explicating H3N2 spreading during three epidemic seasons.

To be able to reconstruct Bayesian coalescent skygrid \citep{minin2008smooth} plot, we exploit Tracer to exact past population dynamics from both MCMC log and tree log produced by a BEAST \citep{drummond2007beast} run. 
The HKY + $\Gamma_4$ model, strict molecular clock, and Bayesian skygrid tree prior was respectively setup in the run, where a grid of 50 intervals over 5 years estimating 10 population sizes by year was constructed in the skygrid prior.

%What type of dynamics does the H3N2 skyride plot suggest?
According to Tracer's skygrid plot in Figure~\ref{fig:flu}.a, the changing patterns of genetic diversity in viral isolates from New York state clearly reveal the seasonal dynamics of influenza A in individual temperate populations. The absolute amount of genetic diversity, even at seasonal peaks, is small compared to other rapidly evolving viruses that infect far fewer people, suggesting that strong natural selection, in addition to periodic bottlenecks, reduces the level of diversity that co-circulates at any time. Simulations demonstrate that this reconstruction of genetic diversity is robust to the sampling protocol. %TODO copied from (Rambaut et al., 2008)

\begin{figure}[ht]
%\includegraphics[width=.5\textwidth]{./figures/fluBSP.pdf}  
\subfigure[Bayesian coalescent skygrid]{\includegraphics[width = .23\textwidth]{./figures/fig21.png}}
\subfigure[Lineages through time]{\includegraphics[width = .23\textwidth]{./figures/fluLTT.pdf}}
\caption{(a) Bayesian coalescent skygrid plot and (b) lineages through time for the Influenza A/H3N2 virus data.}
\label{fig:flu}
\end{figure}

Lineages Through Time in Figure~\ref{fig:flu}.b, similar curve

diversification has been gradually developed before 2003's break-out, but it also looks stopped after 2004.

@@

%\begin{figure}[H]
%\includegraphics[width=.5\textwidth]{./figures/comp-95HPD.pdf}  
%\caption{The comparison of  95\% HPD intervals from multi-trace}
%\label{fig:comp95HPD}
%\end{figure}

%\begin{figure}[H]
%\includegraphics[width=.5\textwidth]{./figures/jointPrInt.png}  
%\caption{The Joint probability table of two integer traces}
%\label{fig:int:jointpr}
%\end{figure}
